---
title: "HP-Exploratory analysis"
author: "preethi"
date: "11/23/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Loaded the required packages

```{r load packages, results='hide',message=FALSE,error=FALSE}
library(googledrive)
library(lubridate)
library(ggplot2)
library(dplyr)
library(scales)
library(reshape2)
```

Download file and read it

```{r downloadfile, results='hide',message=FALSE,error=FALSE}
temp <- tempfile(fileext = ".zip")
dl <- drive_download(
  as_id("https://drive.google.com/open?id=1dyKTsCDegBCDDBDVJRJAf_JIUxd8YcET"), path = temp, overwrite = TRUE)
out <- unzip(temp, exdir = tempdir())
expenditure <- read.csv(out, header = TRUE, stringsAsFactors = FALSE)
unlink(temp)

budget <- read.csv("hoa_wise_prep_data.csv", header = TRUE, stringsAsFactors = FALSE)
```
I cannot seem to download and extract the budget file from the drive through R so I have downloaded it and saved it locally

##temp <- tempfile(fileext = ".zip")
##dl <- drive_download(
  as_id("https://drive.google.com/open?id=1LQNEV3vQDI3nkofVwADHrTZdPIh29oAg"), path = temp, overwrite = TRUE)
##out1 <- unzip(temp, exdir = tempdir())
##budget <- read.csv(out1, header = TRUE, stringsAsFactors = FALSE)

Convert the format of the date from factor to date format

```{r time, results='hide',message=FALSE,error=FALSE}
## change date format

expenditure$TRANSDATE <- as.Date(expenditure$TRANSDATE, format = "%Y-%m-%d")
budget$date <- as.Date(budget$date, format = "%Y-%m-%d")
```

Add a month_year column to sort by month_year


## I am going to assume that the revised and sanctioned data is in 100000's and do the following. 

```{r process}
budget$Revisedlakh <- budget$REVISED *100000
budget$SanctionedLakh <- budget$SANCTION * 100000

```

##budget and expenditure data is then grouped by month_year

```{r month_year}
## expenditure data grouped by month and netpayment

Sumpayment <- expenditure %>% mutate(month_year = format(TRANSDATE, "%Y-%m")) %>% group_by(month_year) %>% summarise(total = sum(NETPAYMENT, na.rm = TRUE))

BudgetSum <- budget %>% mutate(month_year = format(date, "%Y-%m")) %>%
group_by(month_year) %>% summarise_at(c("Revisedlakh","SanctionedLakh"), sum, na.rm = TRUE)

## budget data grouped by medical budget
medical_budget <- budget %>% mutate(month_year = format(date, "%Y-%m")) %>%
group_by(month_year) %>% filter(major == 2210) 

medical_budget_month <- budget %>% mutate(month_year = format(date, "%Y-%m")) %>%
group_by(month_year) %>% filter(major == 2210) %>% summarise_at(c("Revisedlakh","SanctionedLakh"), sum, na.rm = TRUE)

medical_exp_distric <- expenditure %>% mutate(month_year = format(TRANSDATE, "%Y-%m")) %>%
group_by(month_year) %>% filter(major == 2210) 

medical_expenditure <- expenditure %>% mutate(month_year = format(TRANSDATE, "%Y-%m")) %>% group_by(month_year) %>%
filter(major == 2210) %>% summarise(total = sum(NETPAYMENT, na.rm = TRUE))


```

Melt the data so we can visualize the revised and sanctioned estimates as variables

```{r mutate}

BudgetSum.long <- melt(BudgetSum, id = "month_year", measure = c("Revisedlakh","SanctionedLakh"))

medical_budget.long <- melt(medical_budget_month, id = "month_year", measure = c("Revisedlakh","SanctionedLakh"))
```


```{r mutate}

## plot of total expenditure monthly for state

plot1 <- ggplot (data = Sumpayment, aes(x = month_year, y=total, group = 1)) + geom_line(color = "red", size=1)+ geom_point(size=3)+scale_y_continuous(labels = scales::comma)+theme (axis.text.x = element_text(angle = 90, hjust = 1))+ ggtitle("Plot of  monthly netpayment over 2018-2019 in HP") + xlab("time(month-year)") + ylab("payment in Rs")

print (plot1)

##plot of monthly budget data

plot2a <- ggplot (data = BudgetSum.long,aes(x = month_year, y=value, color = variable, group = variable)) + geom_line(size=1)+ geom_point(size=3)+scale_y_continuous(labels = scales::comma)+ theme(axis.text.x = element_text(angle = 90, hjust = 1))+ ggtitle("Plot of monthly budget (Revised & Sanctioned) over 2018-2019 in HP") + xlab("time(month-year)") + ylab("payment in Rs")

print (plot2a)

##plot of budget and expenditure data

plot3 <- ggplot() + geom_line(data = Sumpayment, aes(x = month_year, y=total, group = 1), color = "red", size =1)+ geom_point(size = 3)+ 
  geom_line(data = BudgetSum.long, aes(x= month_year,y=value,colour = variable, group = variable),size =1) + 
geom_point(size=3)+
  scale_y_continuous(labels = scales::comma)+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+ 
  ggtitle("Plot of monthly budget (Revised & Sanctioned) & expenditure over 2018-2019 in HP") + xlab("time(month-year)") + ylab("payment in Rs")

print(plot3)

##plot of monthly medical budget estimates (revised and sanctioned)
plot4a <- ggplot(data = medical_budget.long, aes(x= month_year, y=value, color=variable, group = variable))+ geom_line(size = 1)+ scale_y_continuous(labels = scales::comma)+ geom_point(size=3)+ theme(axis.text.x = element_text(angle = 90, hjust = 1))+ ggtitle("Plot of monthly medical budget (Revised & Sanctioned) over 2018-2019 in HP") + xlab("time(month-year)") + ylab("payment in Rs")

print(plot4a)

##plot of monthly medical expenditures
plot5 <- ggplot (data = medical_expenditure, aes(x = month_year, y=total, group = 1)) + geom_line(color = "red", size=1)+ geom_point(size=3)+scale_y_continuous(labels = scales::comma)+theme (axis.text.x = element_text(angle = 90, hjust = 1))+ ggtitle("Plot of  monthly medical expenditure over 2018-2019 in HP") + xlab("time(month-year)") + ylab("payment in Rs")

print (plot5)

```

For plot 1: January and march seem to have higher values than the other months, we can subset that data to see if we can figure out what is causing this

```{r subset_months}

expenditure_jan <- expenditure[expenditure$TRANSDATE >= "2018-01-01" & expenditure$TRANSDATE >= "2018-01-31",]

```



## Districtwise Spending 

There are supposed to be 12 districts in Himachal. There is a lot more data here. 

## data that is not working yet

Districtspend <- expenditure %>% group_by(District)%>% summarise(Total = sum(NETPAYMENT, na.rm = TRUE))

District_spending_month <- expenditure %>% group_by(month,District,SOE_description) %>% summarise(Total = sum(NETPAYMENT, na.rm = TRUE))

##total expenditure monthly per district

district_plot <- ggplot(data = District_spending_month, aes(x=month,y=Total, group =1))+
  geom_line(color = "darkorchid4") +
   facet_wrap( ~ District, ncol = 7) +
  labs(title = "Total Expenditure by district",
           subtitle = "Data plotted by month",
           y = "total expenditure",
           x = "month") + theme_bw(base_size = 15) +
          scale_y_continuous(labels = scales::comma)

print(district_plot)

##pie chart with district wise spending

barplot <- ggplot(subset(District_spending_month, District == "AMB"), aes(x="",y = Total, fill = SOE_description))+geom_bar(stat="identity", width = 1)

pie <- barplot + coord_polar("y", start=0)

pie



##not clean



            

