---
title: "Proposal-1"
author: "preethi"
date: "11/29/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

 ## Comparison of trends in Himachal Pradesh's Public Health Expenditure with trends in Public health infrasatructure and health indicators

## Background / Introduction

The Indian Constitution charges the state with the task of "improvement of public health" of its people and therefore most of the public health expenditure happens from the state. 

For the year 2015-2016 (according to the National Health Accout Estimates based on the System of Health Account Estimates, SHA-2011), total health expenditure (THE) accounted for about 3.84% of the GDP at 5,28,484 crores, of which the Government Health Expenditure was 1,61,863 crores and accounted for 1.18% of the total GDP and accounting for about 30% of Total Health Expenditure. Of the Goverenment Health Expenditure 64.4% is incurred by the state government. Household Out of Pocket expenditure accounts for about 64% of the THE. 
 
## Demographic information about Himachal
Himachal Pradesh, a Northern Himalayan state which extends over 55,673 km². It has a projected population of 74.35 Lakhs in 2019 and about 90% of this population is Rural. 
The difficult terrain lends itself to complications with the provision of healthcare and in 2005, The central government introduced the National Rural Health Mission (later titled National Health Mission) which it implemented it in 18 states including Himachal Pradesh. 
In the 25 years since its implementation, there have been general improvements observed in Himachal Pradesh's Public health expenditure.
In 2016, the Government's overnment Health Expenditure was 47% of Total Health Expenditure and 1.4% of the state Gross Domestic Product and 6.5% of the General Government Expenditure and it also had the highest per-capita spending of all states at -- however the Out of Pocket expenditure was still about 49.5%. 


However, there have not been many attempts to quanitify the role of public health expenditure on the health of this state over time.
A large body of literature links improved health outcomes to economic growth and poverty reduction. While the link between expenditures and outcomes is never automatic in any country, it is generally positive when expenditures are managed and executed efficiently. Research suggests that increased spending on health, when coupled with good policies and good governance, can promote growth, reduce poverty, and trigger declines in infant, child, and maternal mortality. 
Himachal Pradesh has also seen an increase in the number of functioning PHCs, Sub Centres and CHC's with time. But does this increase translate to Health Indicators. 

The aim of this study would be to map trends in Public Health Expenditure data and compare it with changes in Public Health Infrastructure and estimators of health over time and answer the question "Is the public health expenditure in Himachal Pradesh translating into actual improvements in Public Health Infrastructure and Health Indicators".
Public health policy and expenditure needs to change and adapt to changes in the Public Health landscape.The study would give interesting information and will be useful toward shaping more useful and efficient health policy.

Objectives/ Endpoints

This study aims to look at three different sources of data and 

a) observe trends in Rural Health infrastructure with changes in Public health expenditure. Rural health infrastructure endpoints could be numbers of Primary Healthcare Centres, CHC's and Subcentres, as well as Healthcare manpower in these centres as well as changes to total number of beds.

b) observe trends in health indices with changing public health expenditure. Indices to observe will be Infant Mortality Rate, Under 5 Mortality rate, Maternity Care, Family planning methods, delivery care, vaccinations and anemia amongst children and adults.


Methodology 

This study will be consolidating time series data from three different sources to look for changes with time and associations between them 

We will look at the state budget data especially public health expenditures over time, Rural Health statistics data for changes in Rural Health Infrastructre and lastly we will see trends in health indicators 

We will use the following sources

## State Budget Data

- To observe (state-wide and district-specific) changes in Public health expenditure over time 
as well as changes in the component-wise break up of public health expenditure over time.

There seem to be differing opinions with regard to how public health expenditure is calculated. 

For the purpose of this study we will consider public health expenditure as revenues and capital expenditures on the
budget major heads “Medical and Public Health” and “Family Welfare”.

  -In this case, the relevant Major Heads are 2210 – Revenue Expenditure on Medical and Public Health, 2211 - Revenue Expenditure on Family Welfare, 4210 - Capital Expenditure on Medical and Public Health and 4211- Capital Expenditure on Family Welfare.
  
More in detail
  
The list of Major, Sub-Major, Minor heads for PHC's CHC's and Sub-centres are  

2210-03-103 for PHCs  

2210-03-104 for CHCs  

2211-00-101 for Sub-Centres  

4210-03-103 for capital expenditure on PHCs  
and  

4210-03-104 for capital expenditure on Sub-Centres.


Here I have outlined the Total state expenditure for 2018-2019 and the Public health Component for 2018-2019. As well as the total state revised budget and the public health component for 2018-2019


```{r state_budget_data, echo=FALSE}

library(lubridate)
library(ggplot2)
library(dplyr)
library(scales)
library(reshape2)

budget <- read.csv ("hoa_wise_prep_data.csv", header = TRUE, stringsAsFactors = FALSE) ##read budget data

expenditure <- read.csv("district_treasury_spending_expenditure.csv",header = TRUE, stringsAsFactors = FALSE) ## read expenditure data

budget$date <- as.Date(budget$date, format = "%Y-%m-%d") ##change budget date format

expenditure$TRANSDATE <- as.Date(expenditure$TRANSDATE, format = "%Y-%m-%d") ##change expenditure date format

expenditure <- expenditure %>% mutate(month_year = format(TRANSDATE, "%Y-%m")) ##create a new column with month-year

Monthly_Expenditure <- expenditure %>% group_by(month_year) %>% summarise(Total_Monthly_exp = sum(NETPAYMENT, na.rm = TRUE)) ##create new dataframe with monthly total expenditure

Budget_monthly<- budget %>% mutate(month_year = format(date, "%Y-%m")) %>%
group_by(month_year) %>% summarise_at(c("REVISED","SANCTION"), sum, na.rm = TRUE) ##add column to budget with mon-year and then summarize data by month for both revised and sanctioned budget

colnames(Budget_monthly) <- c("month_year", "Revised_total_budget", "Sanctioned_total_budget")

medical_budget_month <- budget %>% mutate(month_year = format(date, "%Y-%m")) %>%
group_by(month_year) %>% filter(major == 2210,2211,4210,4211) %>% summarise_at(c("REVISED","SANCTION"), sum, na.rm = TRUE)  ## create a new d.f with the 4 major heads and summarize by month and revised and sanctioned data

medical_expenditure <- expenditure %>% mutate(month_year = format(TRANSDATE, "%Y-%m")) %>% group_by(month_year) %>%
filter(major == 2210,2211,4210,4211) %>% summarise(Medical_exp= sum(NETPAYMENT, na.rm = TRUE)) ## do the same with expenditure data

colnames(medical_budget_month) <- c("month_year", "Revised_medical_budget", "Sanctioned_medical_budget")

expenditure_medical_total <- merge(medical_expenditure,Monthly_Expenditure,by="month_year") ##merge the two datasets

budget_medical_total <- merge(medical_budget_month,Budget_monthly, by = "month_year") ##same as above

budget_medical_total_long <- melt(budget_medical_total, id = "month_year", measure = c("Revised_total_budget", "Sanctioned_total_budget","Revised_medical_budget", "Sanctioned_medical_budget")) ##melt the data for easy plottage

expenditure_medical_total_long <- melt(expenditure_medical_total, id = "month_year", measure = c("Total_Monthly_exp","Medical_exp"))
##same as above

plot1 <- ggplot (data = expenditure_medical_total_long,aes(x = month_year, y=value, color = variable, group = variable)) + geom_line(size=1)+ geom_point(size=3)+scale_y_continuous(labels = scales::comma)+ theme(axis.text.x = element_text(angle = 90, hjust = 1))+ ggtitle("Plot of monthly expenditure and public health expenditure in 2018-2019 in HP") + xlab("time(month-year)") + ylab("Expenditure in Rs")

print (plot1)

plot2 <- ggplot (data = budget_medical_total_long,aes(x = month_year, y=value, color = variable, group = variable)) + geom_line(size=1)+ geom_point(size=3)+scale_y_continuous(labels = scales::comma)+ theme(axis.text.x = element_text(angle = 90, hjust = 1))+ ggtitle("Plot of monthly budget and public health budget (sanctioned and revised) in 2018-2019 in HP") + xlab("time(month-year)") + ylab("Budget in Rs")

print (plot2)

```

## Rural Health Statistics Data

- This data is collected by the Ministry of Health and Family Welfare and put out by the National Health Mission (NHM) previously called National Rural Health Mission (NRHM) every year from 2005 till 2018.

- It provides data of Rural Health Statistics. Rural Health Statistics data usually contains Infrastructural information and building positions of Sub Centres, Primary Health Centres (PHCs) and Community Health Centres (CHCs). It also contains state wise number of Sub Divisional Hospital, District Hospital & Mobile Medical Units. As well as state-wise information of Specialist Doctors, Health Workers, Auxiliary Mid wife Nurses (ANM), Nursing Staff, Health Assistants, Radiographers, Pharmacists. It includes information of staff in position, requirement of staffs and shortfall.

- To track trends in Rural Health infrastructure as a factor of time and changes in public health expenditures (numbers of PHCs, CHCs, Sub-centres as well as personel and hospital standards). 
This data may not be available in a district-wise manner for most years.

```{r district_data, echo = FALSE}
district_exp <- read.csv("districtwise_spend.csv", header = TRUE)
district_exp$district <- district_exp$ï..district
district_exp$Population <- as.numeric(district_exp$Population)

plot3 <- ggplot (data = district_exp, 
                 aes(x = Public_Health_Expenditure,y=Number_PHC, group = district, size=Population))+
                  geom_point(alpha = 0.7)+scale_x_continuous(labels = scales::comma)+ 
                  scale_size(range = c(.1, 24), name="Population")+ 
                  theme (axis.text.x = element_text(angle = 90, hjust = 1))+ 
                  ggtitle("Plot of Distric-wise public health expenditure and number of hospitals") +
                  xlab("District wise expenditure in Rs") + ylab("Number of Primary Health Care Centres")
plot3
```

(Show a district-wise graph for 2018 with public health expenditure in the x-axis,numbers of PHC's, Sub-centres and CHC's in x axis and population size as the ize of the circle)

- We can also look at the other endpoints such as health care workers in these centres. 

A prelimnary look at this data suggests that while numbers of PHC's, CHC's and Sub-Centres have increased, there maybe a need for Health Workers as the vacancies and shortfalls indicate. 

(Show bar graphs of some infrastructure changes)

## National Family Health Survey

- This is a large-scale, multi-round survey conducted in a representative sample of households throughout India. This is also conducted by the Ministry of Health and Family Welfare

- There are four surveys available at the moment with data available from 1992-2015

- The latest survey NFHS 2014 outlines 114 health outcome indicators.  

- This study focuses on the relationship between some of the health outcome indicators (specifically vaccination and Infant Mortality) from NFHS-3,NFHS-4 and NFHS-5(the first phase to be released by January 2020) with corresponding changes in public health expenditures.

- Just a prelimnary look at this data shows that while Himachal Pradesh performs well on many of these indicators, vaccination rates have decreased steadily since 2005. 

(Maybe include line graph of vaccination data)

- One of the outcomes of this data analysis could be to consolidate data from these data sets to arrive at possible reasons for this decrease in vaccination rates


...

TIMEFRAMES
STRENGTHS AND WEAKNESSES OF THE STUDY
PUBLIC HEALTH SIGNIFICANCE
BUDGET AND MOTIVATION
REFERENCES
APPENDICES
  - APPENDIX 1: QUESTIONNAIRE
  - APPENDIX 2: PATIENT LIST


```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
